---
title: "Birth Data Report"
author: "Allison Young"
date: "September 22, 2018"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#Load libraries
library(readr)
library(ggplot2)
library(lattice)
library(lubridate)
library(openxlsx)
library(tinytex)


#Working Directory and files
dir <- "C:/Users/ayoung/Desktop/Duke/Git Repositories/Modeling-IDS-702/"  ### INSERT
folder = "Birth Data Analysis"  ### INSERT
fname = "smoking.csv" ### INSERT

# Load Functions
source("../Functions/loadfile.R")
source("../Functions/exploredata.r")
source("../Functions/pullfname.R")
source("../Functions/textsummary.R")

#Load Data
births <- loadfile(dir,folder,fname)
```

### Printed Summary of File
```{r}
textsummary(fname,births)
```

### Explored Data 
Explored diminsions of the file and summaries of variables.
```{r, echo=FALSE}
exploredata(births)
```

### Predictor Variable Transformations and Labeling
-Changed bwt.oz to bwt, mean centered date variable and re-visited converting date variables to strings for interpretation. 
-Created a new column, "mracetxt" to hold the new race variable names, "White" (categories 0-5), "Mexican", "Black", "Asian", and "Mix".
-Created new label columns for mother education and mother income.
-Plotted histograms and box plots of x variables for both continuous and categorical variables.
```{r, include=FALSE}
##Birthweight (continuous, leave as cont, but maybe also do LBW catigorization if time)
births$bwt <- births$`bwt.oz`  #make name easier to use


## Date of Birth (date as a number, convert to date)
births$dateCent <- births$date- mean(births$date)
births$dateMod <- convertToDate(births$date)
# Mother Race (Categorical, needed to condense number of categoires)
births$mracetxt[births$mrace== 0] <- "White"
births$mracetxt[births$mrace== 1] <- "White"
births$mracetxt[births$mrace== 2] <- "White"
births$mracetxt[births$mrace== 3] <- "White"
births$mracetxt[births$mrace== 4] <- "White"
births$mracetxt[births$mrace== 5] <- "White"
births$mracetxt[births$mrace== 6] <- "Mexican"
births$mracetxt[births$mrace== 7] <- "Black"
births$mracetxt[births$mrace== 8] <- "Asian"
births$mracetxt[births$mrace== 9] <- "Mix"
births$mracetxt[births$mrace== 99] <- NA

#Mother Race
births_plotRace <- ggplot(data = births) + geom_bar(mapping = aes(x=mracetxt))
births_plotRace

#Smoking Status (Boolean, Primary Predictor)
births$smoketxt[births$smoke== 0] <- "Non-Smoker"
births$smoketxt[births$smoke== 1] <- "Smoker"
births_plotSmoke <- ggplot(data = births) + geom_bar(mapping = aes(x=smoketxt))
births_plotSmoke

#Parity (Continuous, test if linear assoc relevant to model)
births$parityCent <-  births$parity- mean(births$parity)
births_plotParity <- hist(births$parity)
ordered(unique(births$parity))
#Levels: 0 < 1 < 2 < 3 < 4 < 5 < 6 < 7 < 8 < 9 < 10 < 11
births_plotParity

#Mother's Age (Continuous, could group into teen/non, or linear assoc)
# Should mean center
births_plotMAge <- hist(births$mage)
#ordered(unique(births$mage))
#Levels: 15 < 17 < 18 < 19 < 20 < 21 < 22 < 23 < 24 < 25 < 26 < 27 < 28 < 29 < ... < 45
births$mageCent <- births$mage- mean(births$mage)
births_plotMAge

#Mother's Education (Categorical, check linear association (or Factor in few groups))
births_plotMEd <- hist(births$med)
ordered(unique(births$med))
#Levels: 0 < 1 < 2 < 3 < 4 < 5 < 7
births$medtxt[births$med== 0] <- "Less than 8th Grade"
births$medtxt[births$med== 1] <- "8th to 12th, No Grad"
births$medtxt[births$med== 2] <- "HS Grad Only"
births$medtxt[births$med== 3] <- "HS Grad and Trade"
births$medtxt[births$med== 4] <- "HS Grade and Some College"
births$medtxt[births$med== 5] <- "College Grad"
births$medtxt[births$med== 6] <- "Trade School (HSG Unknown)"
births$medtxt[births$med== 7] <- "Trade School (HSG Unknown)"
births_plotMEd

#Mother's Height (Continuous, test if relevant to model)
births_plotMHt <- hist(births$mht)
ordered(unique(births$mht))
#Levels: 0 < 1 < 2 < 3 < 4 < 5 < 7
births$mhtCent <- births$mht- mean(births$mht)
births_plotMHt

#Mother's Weight (Continuous, test if relevant to model)
births_plotMpregWt <- hist(births$mpregwt)
ordered(unique(births$mpregwt))
births$mpregwtCent <- births$mpregwt- mean(births$mpregwt)
births_plotMpregWt

#Monther's Income (categorical, group into 3 or so categories based on linear analysis)
births_plotInc <- hist(births$inc)
ordered(unique(births$inc))
#Levels: 0 < 1 < 2 < 3 < 4 < 5 < 6 < 7 < 8 < 9
births$inctxt[births$inc== 0] <- "Under 2500"
births$inctxt[births$inc== 1] <- "2500-4999"
births$inctxt[births$inc== 2] <- "5000-7499"
births$inctxt[births$inc== 3] <- "7500-9999"
births$inctxt[births$inc== 4] <- "10000-12499"
births$inctxt[births$inc== 5] <- "12500-14999"
births$inctxt[births$inc== 6] <- "15000-17499"
births$inctxt[births$inc== 7] <- "17500-19999"
births$inctxt[births$inc== 8] <- "20000-22499"
births$inctxt[births$inc== 9] <- "22500+"
births_plotInc

##Check Added Variables
#head(births$bwt)
#head(births$smoketxt)
#head(births$medtxt)
#head(births$dateMod)
#head(births$mageCent)
#head(births$mhtCent)
#head(births$mpregwtCent)

#Smoking Status
#sample size
n = nrow(births)

#create series of text labels for smoking status
births$smokeY = rep(0, n)
births$smokeY[births$smoketxt == "Smoker"] = 1
births$smokeN = rep(0, n)
births$smokeN[births$smoketxt == "Non-Smoker"] = 1

#create series of text labels for mother income variable
births$inctxt[births$inc== 0] <- "Under 2500"
births$inctxt[births$inc== 1] <- "2500-4999"
births$inctxt[births$inc== 2] <- "5000-7499"
births$inctxt[births$inc== 3] <- "7500-9999"
births$inctxt[births$inc== 4] <- "10000-12499"
births$inctxt[births$inc== 5] <- "12500-14999"
births$inctxt[births$inc== 6] <- "15000-17499"
births$inctxt[births$inc== 7] <- "17500-19999"
births$inctxt[births$inc== 8] <- "20000-22499"
births$inctxt[births$inc== 9] <- "22500+"
```

## II. Exploring Relationships
Outcome Variable: Birthweight (bwt)

First, I mapped a series of plots between birthweights and predictor variables, and looked at the relationship between birthweight and smoking. When looking at birthweight and smoking by race, there appeared to potentially be an interaction by race as well as a clear difference between some races, such as between black and white babies.

```{r fig1, fig.height = 5, fig.width = 6, fig.align = "center", echo=FALSE}
boxplot(bwt~smoketxt, data=births, main= "Birthweight by Smoking Status",xlab="Smoking Status", ylab="Birthweight") #Clearly there is an association, now add other variables
bwplot(bwt~as.factor(smoke)|mracetxt, data=births, main="Relationship between Birthweight and Smoking Status by Race", xlab= "Smoking Status", ylab = "Birthweight")
```

## Initial Modeling

After considering inital plots and a few initial models, I settled on the first model to plot out, and check residuals for, found below.


### Model 1 (with interaction)
Baseline Predictor Factors:
Smoker= Non Smoker,
Race= White Mother,
Mother Education = HS Grad Only,
Income = $10,000 - $12,499

Model Variables:
Mother Age,
Mother Height,
Mother Pregnant Weight,
Parity,
Date of Birth

Interaction:
Mother Race(As Factor, White) * Smoke(As Factor, Smoker)

(see end of document for model output for Model 1, as this was the model I settled on)
```{r}
Mod1 <- lm(bwt~mageCent + mhtCent + mpregwtCent + dateCent+ parityCent + relevel(as.factor(medtxt), ref = "HS Grad Only")+ relevel(as.factor(inctxt), ref="10000-12499") +  relevel(as.factor(smoke), ref="0") * relevel(as.factor(mracetxt), ref="White"), data = births)
#summary(Mod1)
```

####Model 1 Residuals by Predictor Variables
The residuals for this model look pretty good.
```{r fig2, fig.height = 15, fig.width = 15, fig.align = "center", echo=FALSE}
par(mfrow=c(3,3))
plot(y=Mod1$residuals, x=births$mage, xlab= "Mother Age")
abline(0,0)
plot(y=Mod1$residuals, x=births$mht, xlab= "Mother Height")
abline(0,0)
plot(y=Mod1$residuals, x=births$mpregwt, xlab= "Mother Pregnant Weight")
abline(0,0)
plot(y=Mod1$residuals, x=births$date, xlab= "Date of Birth")
abline(0,0)
plot(y=Mod1$residuals, x=births$med, xlab= "Mother Education")
abline(0,0)
plot(y=Mod1$residuals, x=births$inc, xlab= "Mother Income")
abline(0,0)
plot(y=Mod1$residuals, x=births$parity, xlab= "Parity")
abline(0,0)
plot(y=Mod1$residuals, x=births$smoke, xlab= "Smoking Status")
abline(0,0)
plot(y=Mod1$residuals, x=births$mrace, xlab= "Mother Race")
abline(0,0)
```

###Model 2 (without interaction)
I also created a second model without interaction and summarized it as well.

Baseline Predictor Factors:
Smoker= Non Smoker,
Race= White Mother,
Mother Education = HS Grad Only,
Income = $10,000 - $12,499 

Model Variables:
Mother Age,
Mother Height,
Mother Pregnant Weight,
Parity,
Date of Birth

No Interaction

```{r}
Mod2 <- lm(bwt~mageCent + mhtCent + mpregwtCent + dateCent+ parityCent + relevel(as.factor(medtxt), ref = "HS Grad Only")+ relevel(as.factor(inctxt), ref="10000-12499") +  relevel(as.factor(smoke), ref="0") + relevel(as.factor(mracetxt), ref="White"), data = births)
summary(Mod2)
```


#####Model 2 Residuals by Predictor Variable
Plotted Variable residuals for Model 2 are similar to those of Model 1.
```{r fig3, fig.height = 15, fig.width = 15, fig.align = "center", echo=FALSE}
par(mfrow=c(3,3))
plot(y=Mod2$residuals, x=births$mage, xlab= "Mother Age")
abline(0,0)
plot(y=Mod2$residuals, x=births$mht, xlab= "Mother Height")
abline(0,0)
plot(y=Mod2$residuals, x=births$mpregwt, xlab= "Mother Pregnant Weight")
abline(0,0)
plot(y=Mod2$residuals, x=births$date, xlab= "Date of Birth")
abline(0,0)
plot(y=Mod2$residuals, x=births$med, xlab= "Mother Education")
abline(0,0)
plot(y=Mod2$residuals, x=births$inc, xlab= "Mother Income")
abline(0,0)
plot(y=Mod2$residuals, x=births$parity, xlab= "Parity")
abline(0,0)
plot(y=Mod2$residuals, x=births$smoke, xlab= "Smoking Status")
abline(0,0)
plot(y=Mod2$residuals, x=births$mrace,, xlab= "Mother Race")
abline(0,0)
```

## Refining the Model
### Are these two models different? Nested F-test
Next, I did a comparison using ANOVA to see if the interaction I included on the first test makes a difference. Results showed that while there appears to be an interaction by race, the p value is too high to say they are signifcant ( 0.1637). [NOTE: The fact that this interaction is not significant is also evident by the large p values of the coefficient estimates for the interaction between variables of mothers race and smoking status in the first model, listed in the final interpretation of the model]
```{r}
anova(Mod1,Mod2)
```

###Multicollinearity
I also looked at whether parity was highly correlated with mother's age, and if so, if it could be removed from the model. Results showed that while there is a correlation (.52), it is not high enough to remove parity from the model.
```{r}
cor(births$mage,births$parity)
```

After these two additional explorations, the best R^2 I was able to achieve was 0.175. I have achieved is I feel good interpreting my selected model. 

## Final Model Interpretation

The Final Model I selected is:

Mod1 <- lm(bwt~mageCent + mhtCent + mpregwtCent + dateCent+ parityCent + relevel(as.factor(medtxt), ref = "HS Grad Only")+ relevel(as.factor(inctxt), ref="10000-12499") +  relevel(as.factor(smoke), ref="0") * relevel(as.factor(mracetxt), ref="White"), data = births)

### Important Predictive Factors
From my analysis, the most important predictor factors of birthweight were smoking status, the mother's race, height, and weight, and the date of birth.

#### Mother Height, 
Holding all other factors constant(Non-smoker,White, HS Grad Only, 10-12.5k, mean age, mean weight, mean date of birth, and mean parity), we are 95% confident that the true mean difference in birthweight for an inch increase in mother's height is between CI[0.52, 1.58], with a point estimate of 1.05 oz. 

Therefore, using this model, we can estimate that for every inch increase in the height of a mother during pregnancy, we would expect to see a 1.05 oz increase in the child's birthweight.

With a p value of 0.000119, this predictor is very highly significant (<.0005).


#### Mother Pregnant Weight
Holding all other factors constant(Non-smoker,White, HS Grad Only, 10-12.5k, mean age, mean height, mean date of birth, and mean parity), we are 95% confident that the true mean difference in birthweight for an inch increase in mother's weight is between CI[0.04, 0.17], with a point estimate of 0.11 oz. 

Therefore, using this model, we can estimate that for every pound increase in the weight of a mother during pregnancy, we would expect to see a 0.11 oz increase in the child's birthweight.

With a p value of 0.001210, this predictor is highly significant(<.005).


#### Mother Age
With a p value of 0.781, this predictor is not significant( >.05).


#### Mother Education
With p values >.05, this predictor is not significant.


#### Date of Birth
Holding all other factors constant(Non-smoker,White, HS Grad Only, 10-12.5k, mean weight, mean height, mean age, and mean parity), we are 95% confident that the true mean difference in birthweight for each additional year in date of birth, is between CI[0.73, 8.54], with a point estimate of 4.64 oz. 

Therefore, using this model, we can estimate that for every year increase in the date of birth, we would expect to see a 4.64 oz increase in the child's birthweight.

With a p value of 0.019987, this predictor is significant (<.05).


#### Parity
With a p value of 0.055566, this predictor is not significant (>.05).


#### Smoking Status
Holding all other factors constant(White, HS Grad Only, 10-12.5k, mean age, mean height, mean weight, mean date of birth, and mean parity, we are 95% confident that the true mean difference in birthweight between a smoking and non-smoking mother (White) is between CI[-12.11, -6.74], with a point estimate of -9.427 oz. 

Therefore, using this model, we can estimate the birthweight of a baby born to a nonsmoking Mother(White, HS Grad Only, 10-12.5k): 125.26 oz. And compare that to the estimated birthweight of a mother of the same demographics, who smokes: 115.83 oz.

With a p value of 0.019987, this predictor is significant (<.05).


#### Income of Mother
With p values >.05, this predictor is not significant.


#### Race of Mother
Holding all other factors constant(Non-smoker,HS Grad Only, 10-12.5k, mean age, mean height, mean weight, mean date of birth, and mean parity), we are 95% confident that the true mean difference in birthweight between a mother who is White and mothers who are of other races is between the following: Asian CI[-12.68, 1.58] & pt est (-5.55) , Black CI[-14.40,-6.11] & pt est(-10.26), Mexican CI[-7.28,8.45] & pt est (0.58), Mix CI[-10.17,9.32] & pt est (-0.42).

Therefore, using this model, we can estimate the birthweight of a baby born to a White Mother(other demographics constant)as 125.26 oz. And compare that to the estimated birthweight of a mother of the same demographics, but of different Races: Asian 119.71 oz., Black 115.00 oz., Mexican 125.84 oz., Mix 124.84 oz.

With a p-value of <.0005, this predictor is very highly significant.


#### Interaction between Smoking and Race
Finally, we can take a look at the interaction between race and smoking status as it pertains to our model. To calculate the interaction affect, we would add the difference from baseline based on smoking status + the difference from baseline based on race, and then also add a third variable, which represents the interaction for that race given smoking status.

For example, a mother who is Black and smokes, will see a difference of -17.24 oz, resulting in an estimated birthweight of 108.02 oz. (as compared to a simple sum of Black race and smoker factors, which would be -19.68).

However, a mother who is Mexican and smokes, will see a difference of 5.44 oz, resulting in an estimated birthweight of 130.7.(as compared to a simple sum of Mexican race and smoker factors, which would be -8.85).

Therefore, it appears an interaction does exists because the combination of race and smoker results in a variation in estimated birthweight, as compared to the independent factors alone.

However, with no p values <.05, we can not say any of these interactions are significant.

```{r}
summary(Mod1)
confint(Mod1)
```



