---
title: "Birth Data Report"
author: "Allison Young"
date: "September 22, 2018"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#Load libraries
library(readr)
library(ggplot2)
library(lattice)
library(lubridate)
library(openxlsx)
library(tinytex)
library(arm)
library(pROC)
library(readr)


#Working Directory and files
dir <- "C:/Users/ayoung/Desktop/Duke/Git Repositories/Modeling-IDS-702/"  ### INSERT
folder = "Birth Data Analysis"  ### INSERT
fname = "smoking.csv" ### INSERT

# Load Functions
source("../Functions/loadfile.R")
source("../Functions/exploredata.r")
source("../Functions/pullfname.R")
source("../Functions/textsummary.R")

#Load Data
births <- loadfile(dir,folder,fname)
```

### Printed Summary of File
```{r}
textsummary(fname,births)
```

### Explored Data 
Explored diminsions of the file and summaries of variables.
```{r, echo=FALSE}
exploredata(births)
```

### Predictor Variable Transformations and Labeling
-Changed bwt.oz to bwt, mean centered date variable and re-visited converting date variables to strings for interpretation. 
-Created a new column, "mracetxt" to hold the new race variable names, "White" (categories 0-5), "Mexican", "Black", "Asian", and "Mix".
-Created new label columns for mother education and mother income.
-Plotted histograms and box plots of x variables for both continuous and categorical variables.
-Will need to also change gestation to a boolean variable for logistic regression analysis
```{r, include=FALSE}
##Birthweight (continuous, leave as cont, but maybe also do LBW catigorization if time)
births$bwt <- births$`bwt.oz`  #make name easier to use


## Date of Birth (date as a number, convert to date)
births$dateCent <- births$date- mean(births$date)
births$dateMod <- convertToDate(births$date)
# Mother Race (Categorical, needed to condense number of categoires)
births$mracetxt[births$mrace== 0] <- "White"
births$mracetxt[births$mrace== 1] <- "White"
births$mracetxt[births$mrace== 2] <- "White"
births$mracetxt[births$mrace== 3] <- "White"
births$mracetxt[births$mrace== 4] <- "White"
births$mracetxt[births$mrace== 5] <- "White"
births$mracetxt[births$mrace== 6] <- "Mexican"
births$mracetxt[births$mrace== 7] <- "Black"
births$mracetxt[births$mrace== 8] <- "Asian"
births$mracetxt[births$mrace== 9] <- "Mix"
births$mracetxt[births$mrace== 99] <- NA

#Mother Race
births_plotRace <- ggplot(data = births) + geom_bar(mapping = aes(x=mracetxt))
births_plotRace

#Smoking Status (Boolean, Primary Predictor)
births$smoketxt[births$smoke== 0] <- "Non-Smoker"
births$smoketxt[births$smoke== 1] <- "Smoker"
births_plotSmoke <- ggplot(data = births) + geom_bar(mapping = aes(x=smoketxt))
births_plotSmoke

#Parity (Continuous, test if linear assoc relevant to model)
births$parityCent <-  births$parity- mean(births$parity)
births_plotParity <- hist(births$parity)
ordered(unique(births$parity))
#Levels: 0 < 1 < 2 < 3 < 4 < 5 < 6 < 7 < 8 < 9 < 10 < 11
births_plotParity

#Mother's Age (Continuous, could group into teen/non, or linear assoc)
# Should mean center
births_plotMAge <- hist(births$mage)
#ordered(unique(births$mage))
#Levels: 15 < 17 < 18 < 19 < 20 < 21 < 22 < 23 < 24 < 25 < 26 < 27 < 28 < 29 < ... < 45
births$mageCent <- births$mage- mean(births$mage)
births_plotMAge

#Mother's Education (Categorical, check linear association (or Factor in few groups))
births_plotMEd <- hist(births$med)
ordered(unique(births$med))
#Levels: 0 < 1 < 2 < 3 < 4 < 5 < 7
births$medtxt[births$med== 0] <- "Less than 8th Grade"
births$medtxt[births$med== 1] <- "8th to 12th, No Grad"
births$medtxt[births$med== 2] <- "HS Grad Only"
births$medtxt[births$med== 3] <- "HS Grad and Trade"
births$medtxt[births$med== 4] <- "HS Grade and Some College"
births$medtxt[births$med== 5] <- "College Grad"
births$medtxt[births$med== 6] <- "Trade School (HSG Unknown)"
births$medtxt[births$med== 7] <- "Trade School (HSG Unknown)"
births_plotMEd

#Mother's Height (Continuous, test if relevant to model)
births_plotMHt <- hist(births$mht)
ordered(unique(births$mht))
#Levels: 0 < 1 < 2 < 3 < 4 < 5 < 7
births$mhtCent <- births$mht- mean(births$mht)
births_plotMHt

#Mother's Weight (Continuous, test if relevant to model)
births_plotMpregWt <- hist(births$mpregwt)
ordered(unique(births$mpregwt))
births$mpregwtCent <- births$mpregwt- mean(births$mpregwt)
births_plotMpregWt

#Monther's Income (categorical, group into 3 or so categories based on linear analysis)
births_plotInc <- hist(births$inc)
ordered(unique(births$inc))
#Levels: 0 < 1 < 2 < 3 < 4 < 5 < 6 < 7 < 8 < 9
births$inctxt[births$inc== 0] <- "Under 2500"
births$inctxt[births$inc== 1] <- "2500-4999"
births$inctxt[births$inc== 2] <- "5000-7499"
births$inctxt[births$inc== 3] <- "7500-9999"
births$inctxt[births$inc== 4] <- "10000-12499"
births$inctxt[births$inc== 5] <- "12500-14999"
births$inctxt[births$inc== 6] <- "15000-17499"
births$inctxt[births$inc== 7] <- "17500-19999"
births$inctxt[births$inc== 8] <- "20000-22499"
births$inctxt[births$inc== 9] <- "22500+"
births_plotInc

#Gestation (create boolean variable)
births$gestpreterm[births$gestation < 270] <- "Pre-term Birth"
births$gestpreterm[births$gestation >= 270] <-  "Full Term Birth"
births$gestation[births$gestation <270] <-  1
births$gestation[births$gestation >= 270] <- 0

##Check Added Variables
#head(births$bwt)
#head(births$smoketxt)
#head(births$medtxt)
#head(births$dateMod)
#head(births$mageCent)
#head(births$mhtCent)
#head(births$mpregwtCent)
#head(births$gestpreterm)
#Smoking Status
#sample size
n = nrow(births)

#create series of text labels for smoking status
births$smokeY = rep(0, n)
births$smokeY[births$smoketxt == "Smoker"] = 1
births$smokeN = rep(0, n)
births$smokeN[births$smoketxt == "Non-Smoker"] = 1

#create series of text labels for mother income variable
births$inctxt[births$inc== 0] <- "Under 2500"
births$inctxt[births$inc== 1] <- "2500-4999"
births$inctxt[births$inc== 2] <- "5000-7499"
births$inctxt[births$inc== 3] <- "7500-9999"
births$inctxt[births$inc== 4] <- "10000-12499"
births$inctxt[births$inc== 5] <- "12500-14999"
births$inctxt[births$inc== 6] <- "15000-17499"
births$inctxt[births$inc== 7] <- "17500-19999"
births$inctxt[births$inc== 8] <- "20000-22499"
births$inctxt[births$inc== 9] <- "22500+"
```

## II. Exploring Relationships
Outcome Variable: Pre-term Birth (gestpreterm)

First, I mapped a series of plots between birth term (pre-term or full-term) and predictor variables, and looked at the relationship between birth term and smoking. 

#### Pre- Term Births
```{r fig1, fig.height = 5, fig.width = 6, fig.align = "center", echo=FALSE}
summary(births$gestation)
```

With a mean of .1887 for the modified gestation variable, we know approximately 19% of the births in this data set were preterm.

#### Smoking Status
```{r}
tapply(births$gestation, births$smoketxt, mean)
```

We can also see that there is a difference in preterm outcomes based on the smoking status of mothers, with 22% of mothers who smoke giving birth to preterm babies whereas about 17% of Non-smoking mothers gave birth to preterm babies.

#### Continuous Variables
```{r}
tapply(births$gestation, births$mracetxt, mean)
tapply(births$gestation, births$medtxt, mean)
tapply(births$gestation, births$inctxt, mean)
```
##### Race of Mother
From this data, it appears that Asian mothers may be more likely to have pre-term birth babies, while White mothers may be least likely to have pre-term babies. With an overall percentage of 19% preterm, Black, Mexican, and Asian mothers are all showing rates of pre-term births above the mean. However, the sample is heavily weighted by the largest proportion being White mothers.

##### Education
It appears there is a difference in percent of births that were preterm between those who did or did not graduate from high school, particularly less than 8th grade education. However we must consider sample size as we think about these groupings.

##### Income
In looking at the patterns for income, it appears that there may be a difference for mothers making less than $2500, as compared to other income groupings. However, science tells us that there could be an interaction between income and race when it comes to preterm births. While the higher income bins appear to be relatively average, the means may look different disaggregated by race and/or considering an interaction between income and race.

#### Continuous Variables

```{r}
binnedplot(births$mage, y=births$gestation, xlab = "Mother Age", ylab = "Pre-Term Births", main = "Binned Age and Pre-term birth cases") 

binnedplot(births$mht, y=births$gestation, xlab = "Mother Height", ylab = "Pre-Term Births", main = "Binned Height and Pre-term birth cases") 

binnedplot(births$mpregwt, y=births$gestation, xlab = "Mother Weight", ylab = "Pre-Term Births", main = "Binned Weight and Pre-term birth cases") 

binnedplot(births$date, y=births$gestation, xlab = "Date of Birth", ylab = "Pre-Term Births", main = "Binned Date and Pre-term birth cases") 

binnedplot(births$parity, y=births$gestation, xlab = "Parity", ylab = "Pre-Term Births", main = "Binned Parity and Pre-term birth cases") 
```
##### Mother's Age
While pretty evenly distributed, there appears to be a bit of a parabola shape with the younger and older mothers showing more cases.

##### Mothers Height
There don't appear to be any strong patterns with Mother's height, but potentially a negative linear trend.

##### Mother's Pregnant Weight
There are some odd patterns in Mother's weight, which may be due to sample size within the bins.

##### Date of Birth
There doesn't seem to be much of any pattern with date of birth, and science knowlege probably notes this as a less relevant predictor. 

##### Parity
There are not a lot of bins, but there appears to potentiall be a slight positive linear trend with parity. However it may be worth looking at whether or not there is multicollarity between parity and mothers age.However I did this analysis in the last assignment, and results showed that while there is a correlation (.52), it is not high enough to remove parity from the model.


### Initial Modeling 
```{r}
## take previous analyis model and tweak for new data question
PTMod1 = glm(gestation ~ mageCent + mhtCent+ mpregwtCent + dateCent+ parityCent + relevel(as.factor(medtxt), ref = "HS Grad Only")+ relevel(as.factor(inctxt), ref="10000-12499") +  relevel(as.factor(smoke), ref="0") * relevel(as.factor(mracetxt), ref="White"), data = births, family=binomial)
               
#summary(PTMod1)
roc(births$gestation, fitted(PTMod1), plot=T, legacy.axes=T)

### AOC = .6784
```

```{r}
##add interaction between income and race
PTMod2 = glm(gestation ~ mageCent + mhtCent+mpregwtCent + dateCent+ parityCent + relevel(as.factor(medtxt), ref = "HS Grad Only")+ relevel(as.factor(smoke), ref="0")+ relevel(as.factor(inctxt), ref="10000-12499")*relevel(as.factor(mracetxt), ref="White") , data = births, family=binomial)
#summary(PTMod2)
roc(births$gestation, fitted(PTMod2), plot=T, legacy.axes=T)

## AUC = 0.7165
```

```{r}
## tried a boolean for education, but didn't add much, actually made worse.so keep as levels. (AOC =.6863)
births$medbool[births$med == 0] <- 0
births$medbool[births$med > 0] <- 1

PTMod3 = glm(gestation ~ mageCent + mhtCent + mpregwtCent + dateCent+ parityCent + relevel(as.factor(medbool), ref = 1)+ relevel(as.factor(inctxt), ref="10000-12499")*relevel(as.factor(mracetxt), ref="White") +  relevel(as.factor(smoke), ref="0") * relevel(as.factor(mracetxt), ref="White"), data = births, family=binomial)
               
#summary(PTMod3)
roc(births$gestation, fitted(PTMod3), plot=T, legacy.axes=T)

### AOC = .6863

```

```{r}
#Looked at queadratic for preg weight and mage, only weight really helped. So, I also looked at log of preg weight , all with little result. To keep interpretation as simple as possible, will leave these features out.

births$mpregwtCent2 <- births$mpregwtCent^2
PTMod4 = glm(gestation ~ mageCent + mpregwtCent + mpregwtCent2 + mhtCent +dateCent+ parityCent + relevel(as.factor(medtxt), ref = "HS Grad Only")+ relevel(as.factor(inctxt), ref="10000-12499")*relevel(as.factor(mracetxt), ref="White") +  relevel(as.factor(smoke), ref="0") * relevel(as.factor(mracetxt), ref="White"), data = births, family=binomial)

#summary(PTMod4)
roc(births$gestation, fitted(PTMod4), plot=T, legacy.axes=T)

## AUC = 0.7178
```
```{r}
#However, another idea was to categorize age, and then look at interactions with mother age and race
births$magecat[births$mage > 34] <- "Advanced Age Pregnancy"
births$magecat[births$mage <= 34] <- "'Healthy'Child Bearing Age Pregnancy"
births$magecat[births$mage < 20] <- "Teen Pregnancy"

PTMod5 = glm(gestation ~ relevel(as.factor(magecat), ref= "'Healthy'Child Bearing Age Pregnancy") + mhtCent+mpregwtCent + dateCent+ parityCent + relevel(as.factor(medtxt), ref = "HS Grad Only")+ relevel(as.factor(smoke), ref="0")+ relevel(as.factor(inctxt), ref="10000-12499")*relevel(as.factor(mracetxt), ref="White")+ relevel(as.factor(magecat), ref= "'Healthy'Child Bearing Age Pregnancy")*relevel(as.factor(mracetxt), ref="White") , data = births, family=binomial)
#summary(PTMod5)
roc(births$gestation, fitted(PTMod5), plot=T, legacy.axes=T)

# AUC = 0.7205

```


After considering inital plots and a few initial models, I settled on this last model to run diagnostics on because it makes the most sense considering science, interprebility, and area under the curve.

### Model 5 
Baseline Predictor Factors:
Smoker= Non Smoker,
Race= White Mother,
Mother Age = 'Healthy' Child Bearing Age Pregnancy
Mother Education = HS Grad Only,
Income = $10,000 - $12,499

Model Variables:
Mother Height (centered),
Mother Pregnant Weight (centered),
Parity (centered),
Date of Birth (centered)

Interactions:
Mother Race(As Factor, White) * Mother Income(as Factor, $10000-$12499)

(see end of document for model output for Model 2, as this was the model I settled on)


####Model 5 Residuals 
The residuals for this model look pretty good. The binned plots look evenly distributed, and the mean tables comparing residuals to each of the categorical variables have very low numbers.
```{r fig2, fig.height = 8, fig.width = 8}
rawresid1 = births$gestation - fitted(PTMod5)


binnedplot(x=births$parityCent, y = rawresid1, xlab = "Parity Centered", ylab = "Residuals", main = "Binned residuals versus Parity Centered")

binnedplot(x=births$mhtCent, y = rawresid1, xlab = "Mother Height Centered", ylab = "Residuals", main = "Binned residuals versus Mother Height Centered")

binnedplot(x=births$dateCent, y = rawresid1, xlab = "Date of Birth Centered", ylab = "Residuals", main = "Binned residuals versus Date of Birth Centered")

binnedplot(x=births$mpregwtCent, y = rawresid1, xlab = "Mother Preg Weight Centered", ylab = "Residuals", main = "Binned residuals versus Mother Preg Weight Centered")

tapply(rawresid1, births$medtxt, mean) 
tapply(rawresid1, births$magecat, mean)
tapply(rawresid1, births$inctxt, mean)
tapply(rawresid1, births$mracetxt,mean)
tapply(rawresid1, births$smoke, mean)
```

```{r Model B Diag ConfM}
#let's do the confusion matrix

threshold = 0.5
table(births$gestation, PTMod5$fitted > threshold)

threshold = 0.58
table(births$gestation, PTMod5$fitted > threshold)
```

After these additional explorations, the best area under the curve I was able to achieve while also feeling interpretable was 0.7205 and I feel good moving on to interpreting my selected model. 

## Final Model Interpretation

The Final Model I selected is as follows:
```{r}

PTMod5 = glm(gestation ~ relevel(as.factor(magecat), ref= "'Healthy'Child Bearing Age Pregnancy") + mhtCent+mpregwtCent + dateCent+ parityCent + relevel(as.factor(medtxt), ref = "HS Grad Only")+ relevel(as.factor(smoke), ref="0")+ relevel(as.factor(inctxt), ref="10000-12499")*relevel(as.factor(mracetxt), ref="White")+ relevel(as.factor(magecat), ref= "'Healthy'Child Bearing Age Pregnancy")*relevel(as.factor(mracetxt), ref="White") , data = births, family=binomial)
summary(PTMod5)
#roc(births$gestation, fitted(PTMod5), plot=T, legacy.axes=T)

```



### Important Predictive Factors
```{r}
#coef(PTMod5)

exp(0.49) # Advanced age
exp(-0.29) # Teen birth
exp(-0.014) #m preg weight
exp(0.85) # less than 8th grade ed
exp(-0.36) # lowest income white
exp(0.63) #lowest income asian

exp(confint.default(PTMod5))
```
#### Interpretation Summary

From my analysis, given all other factors constant, it does not appear that smoking has a significant predictive status for a pre-term birth. 

The most important predictor factors of a preterm birth according to my model were the mother's age, pregnant weight, and income, where the relationship between preterm birth and income varies depending on Mother's race. 

Both younger and older moms were more likely to have a preterm birth, as well as mothers that are low income, particularly low income Asian mothers. 

#### Note to TA

I recognize this homework is incomplete. In an ideal world I would interpret the coefficient and confidence interval values of the odds for the predictor factors listed above in my summary. 

I did not leave enough time to complete the assignment after returning from my fall break travels. I will work to finish this assignment over the weekend (for my own practice), and look forward to get another shot at working through this kind of problem on the next assignment.


