---
title: "Job Training Effectiveness Analysis"
author: "Allison Young"
date: "Octoer 15, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Setup and Load Analysis Functions
Load libraries and functions I've created to help with basic data analysis and dataset preview
## Analysis Set-Up
### Load Data Libraries, Files (INSERT FILE NAMES HERE), and Functions
```{r, include=FALSE}

#Load libraries
library(readr)
library(ggplot2)
library(lattice)
library(lubridate)
library(openxlsx)


#Working Directory and files
dir <- "C:/Users/ayoung/Desktop/Duke/Git Repositories/Modeling-IDS-702/"  ### INSERT
folder = "Job Training, Project 1"  ### INSERT
fname = "lalondedata.txt" ### INSERT

#Load Data
ref <- loadfile(dir,folder,fname)

# Load Functions
source("../Functions/loadfile.R")
source("../Functions/exploredata.r")
source("../Functions/pullfname.R")
source("../Functions/textsummary.R")
```

### Printed Summary of File
```{r}
textsummary(fname,ref)
```
############# (Ctrl+F and Replace file names for rest of analysis) #################

### Explore Data 
```{r}
exploredata(ref)
```

### Variable Notes 
```{r}
#### Categorized properly? Continuous vs Categorical-- may need to factorize

#### Clear outliers? Mins and Max--- leverage? or do we need to exclude from model

#### Science considerations? -- interactions, etc- things you need to keep in mind 


#### Outcome Variable?


```


### Data Cleaning (drop columns, drop rows - create dictionary to run through)
```{r}
#### Formula to drop columns 
#makedicM(VALUES- create easy way---manual)
#dropcol(colhead) OR dropcols(dictionary)

#### Formula to drop rows based on value/s 
#droprow(colhead,value)  OR droprows(dictionary)

#### Formula to add labels for rows
#makedicL(VALUES- create easy way---loop)
#addlabels(dictionary)

```


### Initial look(s) at relationships (dictionary, hist, plot y to each x)
```{r}
### Add any summaries of interest

##### Formula to create a dictionary that can use formula to compare y and x
#### Formula to compare y to each x
#plot(outcome~predictorA + predictorB + predictorC, data=ref)
contp <- makedicM(continuous predictors)
catp <-  makedicM(categorical predictors)

plotcat(ref,outcome,contp,labcol)
plotcont(ref,outcome,catp,labcol)

#plot(y = sesame$postnum, x = sesame$prenum, xlab = "Pretest Numbers", ylab = "Posttest numbers")
#boxplot(postnumb ~ viewenc, data = sesame, xlab = "Encouraged", ylab = "Posttest Numbers")

remember--- cat and cont diff-- so may need to be "plot all cat", and "plot all cont" and do labels that way too
```

### Check for Muliticoliarity
```{r}

#### Formula to look through results of below code, bring back anything over .6 or < -.6 (to flag)
#take a look at correlations among predictors for multicollinearity
## highcol(ref, range) -- returns var1, var2, corr
round(cor(ref), #dec)
```

### Center Continuous Variables
### Plot Residuals
### Measure Leverage

#### Other pipedreams
```{r}
#### Formula that is easy to understand to create any kind of dictionary....is that diff?
```

